---
title: "Polling Simulation: Sampling and the Central Limit Theorem"
---

In this activity, you'll use data from the 2024 Cooperative Election Study (CES) — a large-scale national survey conducted during the 2024 election — to simulate what happens when a news organization tries to estimate public opinion from a poll sample. The full dataset has nearly 55,000 respondents, which we'll treat as our population.

Your goal is to see firsthand how sample size affects the accuracy of estimates, and what the distribution of sample estimates actually looks like — the core idea behind the Central Limit Theorem.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from this URL and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

We'll focus on one question: what share of American adults identify as religiously unaffiliated (atheist, agnostic, or nothing in particular)? This has been a major story in American religion journalism for years.

First, we'll create a new column called `unaffiliated` that is `TRUE` if a respondent's `trad` value is `"Atheist"`, `"Agnostic"`, or `"Nothing in Particular"`, and `FALSE` otherwise. Then calculate the proportion of the full dataset that is unaffiliated. Save this value as `population_proportion`.

```{r}
ces <- ces |>
  mutate(unaffiliated = trad %in% c("Atheist", "Agnostic", "Nothing in Particular"))

proportion <- mean(ces$unaffiliated)
```

That population proportion is the "true" population value that any poll is trying to estimate. Write it down — you'll be comparing everything else to it.

### Task 4: Take a single small sample

Use `set.seed(42)` and then sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of unaffiliated respondents in your sample.

```{r}

set.seed(42) 

sample_50 <- ces |> sample_n(50)


proportion_50 <- mean(sample_50$unaffiliated)

```

How close is your sample estimate to the original proportion? Is it surprising?

The sample is about 4 less than the original proportion. This is not surprising because as we learned through the M'M activity, when you do sampling from the same population the proportions of the samples are going to vary. People who set the set seed something other than 42 got different answers because they basically took a different sample. 

### Task 5: Take a single larger sample

Now sample 500 rows. Set a different seed value, grab the sample and save it as `sample_500`. Calculate the proportion of unaffiliated respondents in this sample.

```{r}
set.seed(42) 

sample_500 <- ces |> sample_n(500)


proportion_500 <- mean(sample_500$unaffiliated)

```

How did the estimate change compared to the sample of 50? What does this suggest about sample size?

When sampling for 500, I got 0.362 which is way closer to the original proportion than when I did a sample of 50. This suggests that doing more samples will get you closer to the actual proportion. So, if you are trying to be as accurate as possible you should do as many samples as you can. 

### Task 6: Simulate many polls

A single sample can get lucky or unlucky. To understand how polling actually works, we need to see the *distribution* of estimates across many samples. The code below draws 1,000 samples of size 50, calculates the proportion of unaffiliated respondents in each, and stores the results.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$unaffiliated)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side — one for the 1,000 samples of size 50 and one for the 1,000 samples of size 500. Add a vertical line at `proportion` to each so you can see how well the samples are centering on the true value.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share Religiously Unaffiliated",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**What shape do both histograms have?** Describe them in plain language. Does it surprise you that this shape emerges from a variable that is simply TRUE or FALSE for each respondent?

It isn't surprising that this shape emerges. For histogram with the sample of 50, the curve looks like a pretty normal bell curve ranging from 0.2 to 0.55 while the sample of 500 has a very very small range of 0.3 to 0.43, it is tighter and more compact. 

**Compare the two histograms.** What is the most visible difference between the n=50 and n=500 distributions? What does that difference tell you about what happens when a poll uses a larger sample?

It doesn't surprise me the the shapes are so different because the sample of 500 is way closer to the red line showing the true population proportion while the sample of 50 is less accurate and has values that are further away from the true proportion. This further demonstrates that sampling more will give you more accurate results that will represent the true proportion. 

**Both distributions are centered near the red line.** What does it mean that 1,000 different random samples — each one drawn independently — tend to cluster around the same value? What would it mean for journalism if that *weren't* the case?

Even though the samples are drawn independently they tend to cluster around the same value because they are drawn from the same population. Just like during class when we did the M'Ms we all got different samples but since they were from the same population our percentages were all close to 0.5. If this wasn't the case, it would be a lot harder to write a story about and the journalist would have to do a lot more digging. 

**A reporter tells you their story is based on a poll of 50 people.** Based on what you've seen in the simulation, what questions would you want to ask before trusting that number? You don't need to use any technical terms — just describe what you observed.

I would want to make sure that they sampled a diverse group of people and didn't just sample all of the same people. So I would want to know who they polled and where the polling took place. I would also ask how their close their samples are to the population mean because as we just learned, sampling more is more accurate but it also can be more expensive. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
