---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")

```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}
ces <- ces |>
  mutate(evangelical = trad %in% c("White Evangelical"))

population_proportion <- mean(ces$evangelical)

```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

The value I got was 0.1, meaning that 10% of the people in the dataset are White Evangelical. This is a lot less than the unaffiliated proportion of about 0.36 from in class. So in comparison, the unaffiliated group makes up a large part of the data set compared to White Evangelicals.  

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42) 

sample_50 <- ces |> sample_n(50)


proportion_50 <- mean(sample_50$evangelical)


```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

The sample proportion was 0.16 which is higher than the population proportion of 0.10. During the in-class activity the sample proportion was lower than the population proportion for the unaffiliated group. This is probably because small samples vary from the true population.  

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}
set.seed(1966) 

sample_500 <- ces |> sample_n(500)


proportion_500 <- mean(sample_500$evangelical)

```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

Yes, when taking a larger sample of 500 the proportion of evangelicals was 0.09 which is very close to the population proportion of 0.10. This is similar to what I saw in class because when I did the sample of 500 in class I got 0.362 which was way closer to the original proportion than when I did a sample of 50.

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share of White Evangelicals",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?

The n=50 histogram isn't as symmetric as the unaffiliated one from class. In class, the distribution was more balances around the center but for white evangelicals the distribution is more uneven. Since white evangelicals only make up 10% of the population this makes sense since the unaffiliated group made up a larger share around 36%. 

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

The n=500 histogram is much more centered around the population proportion compared to the sample of 50 because that sample varied more. But when we increase the sample size to 500 the estimate clusters much closer to the true population proportion. The improvement is larger than what  we saw in class because white evangelicals make up a smaller part of the population. 

**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

Based on my simulations, if there was a poll of 200 respondents there would likely be around 20 white evangelical respondents. This suggests that estimates for smaller subgroups are less reliable because the number of respondents is smaller. 

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

I would want to know how many white evangelical respondents were included in the poll. White Evangelicals only make up 10% of the population so even though the poll sampled 400 people that would include about 40 respondents. I would also want to know how the sample was collected and what sampling method was used because sometimes some methods have a higher risk of error. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS. 
