---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Katherine Audrey Ring"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
moco_health<- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats<-moco_health%>%
  summarize(
    mean_compliance = mean(compliance_score, na.rm = TRUE),
    sd_compliance = median(compliance_score, na.rm = TRUE),
    min_compliance = min(compliance_score, na.rm = TRUE),
    max_compliance = max(compliance_score, na.rm = TRUE)
  )


```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
moco_health |> 
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 2) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score) - sd(compliance_score)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(compliance_score) + sd(compliance_score)), color = "blue", linetype = "dashed", size = 1)+
  labs(
    title = "Distribution of MoCo compliance stats",
    x = "complaince amount",
    y = "Count",
  ) +
  theme_minimal()

```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:** Most of the data is clustered around the upper levels showing that a lot of the restaurants do have a good amount of compliance. The standard deviation lines do show some variation around the mean but the data is still showing that most of the compliance scores are high. Due to this, the most newsworthy thing would be the fact that there is a tail on the left showing that some of the restaurants don't have high compliance scores so I would go to those restaurants to see what was happening there compared to the rest of the restaraunts. 
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
transit_WMATA <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats<- transit_WMATA |>
  summarise(
    mean_bus = mean(bus, na.rm = TRUE),
    sd_bus = sd(bus, na.rm = TRUE),
    mean_rail = mean(rail, na.rm = TRUE),
    sd_rail = sd(rail, na.rm = TRUE)
  )


```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- transit_WMATA |>
  mutate(
  rail_zscore = (rail - mean(rail, na.rm = TRUE)) / sd(rail, na.rm = TRUE),
    bus_zscore = (bus - mean(bus, na.rm = TRUE)) / sd(bus, na.rm = TRUE)
  )
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore)))


```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(1966) 

sample_91 <- transit_WMATA |> sample_n(91) 

mean_bussamp <- mean(sample_91$bus)
sd_bussamp <- sd(sample_91$bus)
mean_railsamp <- mean(sample_91$rail)
sd_railsamp <- sd(sample_91$rail) 
```
The population mean for bus is lower than the sample mean for bus ridership but the population standard deviation for bus is greater than the sample standard deviation for bus. The population mean and the standard deviation for rail riders is lower than the sample mean and the standard deviation for rail riders. This makes sense with random sampling because the sample is not going to give the exact population statistics, the sample was a smaller one so while the numbers are close to the population statistics which is good, random sampling isn't exact. 

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)
The unusual days with very low z-scores appear around the holidays including Christmas, Thanksgiving, and New Years Day. This makes sense because people are less likely to use public transit on major holidays, this could make a story about how much income the public transit isn't bringing in during holidays. Another interesting story would be how abnormal the January 6th z-score is as that was when the capital was stormed. 

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
car_theft <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- car_theft %>% mutate(
  theft_rate = (thefts_2022 + thefts_2023 / population) * 10000
)
```
Per capita rates are more useful than raw counts for this comparison because each maryland county is a different size in population so the raw data is more misleading for journalists than the per capita rate. 

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
median_theft <- md_car_thefts  %>% summarize(median_theft = median(theft_rate, na.rm = TRUE))

total_theft <- md_car_thefts %>%
  summarize(total_thefts = sum(thefts_2022+thefts_2023, na.rm = TRUE))

median_theft <- md_car_thefts  %>% summarize(mean_theft = median(total_theft, na.rm = TRUE))

```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)
I can't figure out how to find the ones above the median however the total car theft is 35820
---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points) 
I would prioritize the standard deviation to see what pools are under preforming. Since the mean is 187 and the standard deviation is 42, any pool that data is within 1-2 sd below the mean could show that there are a low number of people there. This would allow me to see the outliers and what pools really are underpreforming. 


### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)
If i calculate the z-score of all of the pools then i could see what pools have abnormal z-scores which would be from -2 and lower which would show that those pools really are statistically underpreforming. Its more useful because it would be statistically accurate and would show how different the pools compare to each other. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)
I would want to know who they took the sample of and make sure it was truly random and if there was any sampling error. I would want to know what months and days of the week they sampled and if the sampling remained the same accross all twelve pools. I would raise questions about sampling error and how small the sample is of 60 days. While it may get closer to the accurate number, random sampling varies and that would have to be clear. 
---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
